### 2015-01-10 doc/thesis/system.tex
Minor improvements to the text about typing Lua Standard Library and MD5
Minor improvements to the section about types

In this chapter we present the formalization of Typed Lua's type system.
Besides its practical contribution, Typed Lua also has some interesting
contributions to the field of optional type systems for scripting
languages.
They are novel type system features that let Typed Lua cover several Lua idioms
and features, such as the refinement of tables, multiple return values,
and optional parameters.

@@

\begin{figure}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
\multicolumn{3}{c}{\textbf{First-level types}}\\
t ::= & \;\; l & \textit{literal types}\\
& | \; b & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; t \cup t & \textit{disjoint union types}\\
& | \; s \rightarrow r & \textit{function types}\\
& | \; \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed|open|unique} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.t & \textit{recursive types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
l ::= & \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} & \\
b ::= & \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String & \\
k ::= & \;\; l \; | \; b \; | \; \Any & \textit{key types}\\
v ::= & \;\; t \; | \; \Const \; t & \textit{value types}\\ 
\multicolumn{3}{c}{}\\
\multicolumn{3}{c}{\textbf{Second-level types}}\\
s ::= & \;\; \Void & \textit{void type}\\
& | \; t* & \textit{variadic types}\\
& | \; t \times s & \textit{tuple types}\\
r ::= & \;\; s \; | \; s \sqcup r & \textit{return types}\\
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure}

Figure \ref{fig:typelang} presents the abstract syntax of the
Typed Lua types.
Typed Lua splits types into two categories:
\emph{first-level types} and \emph{second-level types}.
First-level types represent first-class Lua values and
second-level types represent tuples of values that appear in 
assignments and function applications.
First-level types include literal types, base types, the type $\Nil$,
the top type $\Value$, the dynamic type $\Any$, the type $\Self$,
union types, function types, table types, recursive types, and
projection types.
Second-level types include the type $\Void$, variadic types,
tuple types, and unions of tuple types.

Literal types $l$ represent the type of literal values.
They can be the boolean values $\False$ and $\True$,
an integer value, a floating point value, or a string value.
We shall see that literal types are important in our treatment of
table types as records.

Typed Lua includes four base types $b$: $\Boolean$, $\Integer$, $\Number$, and $\String$.
The base types $\Boolean$ and $\String$ represent the values that during
run-time Lua tags as \texttt{boolean} and \texttt{string}, respectively.
Lua 5.3 introduced two internal representations to the tag \texttt{number}:
\texttt{integer} for integer numbers and \texttt{float} for real numbers.
Both \texttt{integer} and \texttt{float} values are a subtype of \texttt{number}.
We introduced the base type $\Number$ to represent either the tag
\texttt{number} or \texttt{float} values, and the type $\Integer$ to represent
\texttt{integer} values.
In the next section we will show that $\Integer$ is subtype of $\Number$.
This allows programmers to use \texttt{integer} values in the
place of \texttt{float} values.

The type $\Nil$ represents the type of the value that during run-time
Lua tags as \texttt{nil}.

The type $\Value$ is the top type.
In Section \ref{sec:rules} we will show that this type,
along with variadic types, helps the type system to drop extra values
on assignments and function calls, thus preserving the
semantics of Lua in these cases.

Typed Lua uses the type $\Self$ to represent the \emph{receiver}
in object-oriented method definitions and method calls.
As we mentioned in Chapter \ref{chap:typedlua}, we need the type
$\Self$ to prevent programs from indexing a method without
calling it with the correct receiver.

@@

Function types have the form $s \rightarrow r$ to represent Lua functions,
where $s$ is the type of the parameter list and $r$ is the return type.
Both $s$ and $r$ are second-level types.
In the next paragraphs we explain why we need different representations
for the type of the parameter list and the return type.

Second-level types $s$ are tuples of first-level types that can end
with either an empty tuple or with a variadic type.
Typed Lua needs second-level types because tuples are not first-class
values in Lua, though they may appear on argument passing,
multiple returns, and multiple assignments.
The type $\Void$ is the type of an empty tuple.
A variadic type $t*$ is a generator for a sequence of values
of type $t \cup \Nil$, and it represents the type of a vararg expression.

Second-level types $r$ include the same second-level types introduced
by $s$ plus unions of tuple types.
Typed Lua includes unions of tuples because Lua programs
usually overload the return type of functions to denote error,
as we mentioned in Section \ref{sec:statistics}.
We could not find a nice concrete syntax to represent unions of
tuples that appear in the type of the parameter list, and
that is the reason why we restricted them to appear in the return type only.
The same reason led us to split second-level types into two
different symbols.
We use the symbol $\sqcup$ to represent the union between two
different tuple types.
We have different unions because $\cup$ represents the union
between two first-level types, while $\sqcup$ represents the
union between two tuple types.

Back to first-level types, table types represent the various forms
that Lua tables can take.
The syntactical form of table types is $\{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{tag}$,
where each $k_{i}$ represents the type of a table key,
and each $v_{i}$ represents the type of the value that table keys of type $k_{i}$ map to.
Key types can only be literal types, base types, or the dynamic type.
We made this restriction to the type of the keys because the statistics
that we discussed in Section \ref{sec:statistics} showed that most
of the tables are records, lists, and hashes.
The dynamic type is an option when we need a loose table type.
Value types can be any first-level type, and can optionally include
the $\Const$ type to denote immutable values.

We also use the tags \emph{closed}, \emph{open}, and \emph{unique}
to classify table types.
The tag \emph{unique} represents tables that we know their contents,
and that there is no alias to them.
In particular, the type of the table constructor has this tag.
The tag \emph{open} represents \emph{unique} table types that
have at least one alias.
The tag \emph{closed} represents table types that do not provide
any guarantees about the types not listed in the table type.
In particular, type annotations as well as interfaces and userdata
declarations always describe \emph{closed} table types.
In the next sections we explain in more detail why we need
different table types.

@@

Type Lua includes projection types as a way to project
unions of tuple types into unions of first-level types.
Projection types have the form $\pi_{i}^{x}$, where $x$ is a variable
that stores an union of tuple types in a special type environment,
and $i$ is an index that indicates how to project a first-level
union type from $x$.
For instance, if $x$ is the union of tuple types
$(\Integer \times \Integer) \sqcup (\Nil \times \String)$,
then $\pi_{1}^{x}$ represents the type $\Integer \cup \Nil$
while $\pi_{2}^{x}$ represents the type $\Integer \cup \String$.
In Section \ref{sec:rules} we show in more detail how our type system
uses them as a mechanism for handling unions of tuple types,
when they appear on the right-hand side of the declaration of local variables.
We also show how this feature allows our type system to constrain
the type of a local variable that depends on the type of another local variable.

@@

However, sometimes we want to use \emph{unique} table types to initialize
\emph{closed} table types, as they allow us to initialize optional fields.
To do that, Typed Lua includes a cast expression that allow us
to use an \emph{unique} table type before it becomes \emph{open}.
The rule \textsc{T-CAST} describes this behavior:
\[
\begin{array}{c}
\mylabel{T-CAST}\\
\dfrac{t \subtype \env_{1}(n)}
      {\env_{1} \vdash {<}t{>} \; n:t, \env_{1}[n \mapsto t]}
\end{array}
\]

>>>>

In this chapter we present the formalization of Typed Lua's type system.
Besides its practical contribution, Typed Lua also has some interesting
contributions to the field of optional type systems for scripting
languages.
They are novel type system features that let Typed Lua cover several Lua idioms
and features, such as refinement of tables, multiple return values,
and optional parameters.

@@

\begin{figure}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
t ::= & & \textsc{first-level types:}\\
& \;\; l & \textit{literal types}\\
& | \; b & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; t \cup t & \textit{disjoint union types}\\
& | \; s \rightarrow s & \textit{function types}\\
& | \; \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed|open|unique} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.t & \textit{recursive types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
%\multicolumn{3}{c}{}\\
l ::= & & \textsc{{\small literal types:}}\\
& \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
b ::= & & \textsc{{\small base types:}}\\
& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\
%\multicolumn{3}{c}{}\\
k ::= & & \textsc{{\small key types:}}\\
& \;\; l \; | \; b \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
v ::= & & \textsc{{\small value types:}}\\
& \;\; t \; | \; \Const \; t &\\ 
%\multicolumn{3}{c}{}\\
s ::= & & \textsc{second-level types:}\\
& \;\; p & \textit{tuple types}\\
& | \; s \sqcup s & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
p ::= & & \textsc{{\small tuple types:}}\\
& \;\; \Void & \textit{void type}\\
& | \; t* & \textit{variadic types}\\
& | \; t \times p & \textit{pair types}
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure}

Figure \ref{fig:typelang} presents the abstract syntax of
Typed Lua types.
Typed Lua splits types into two categories:
\emph{first-level types} and \emph{second-level types}.
First-level types represent first-class Lua values and
second-level types represent tuples of values that appear in 
assignments and function applications.
First-level types include literal types, base types, the type $\Nil$,
the top type $\Value$, the dynamic type $\Any$, the type $\Self$,
union types, function types, table types, recursive types, and
projection types.
Second-level types include the type $\Void$, variadic types,
tuple types, and unions of tuple types.
Types are ordered by a subtype relationship that we introduce
in Section \ref{sec:subtyping}, so Lua values may belong to
several distinct types.

Literal types represent the type of literal values.
They can be the boolean values $\False$ and $\True$,
an integer value, a floating point value, or a string value.
We will see that literal types are important in our treatment of
table types as records.

Typed Lua includes four base types: $\Boolean$, $\Integer$, $\Number$, and $\String$.
The base types $\Boolean$ and $\String$ represent the values that during
run-time Lua tags as \texttt{boolean} and \texttt{string}, respectively.
Lua 5.3 introduced two internal representations to the tag \texttt{number}:
\texttt{integer} for integer numbers and \texttt{float} for real numbers.
Lua does automatic promotion of \texttt{integer} values to \texttt{float}
values as needed.
We introduced the base type $\Number$ to represent \texttt{float} values,
and the base type $\Integer$ to represent \texttt{integer} values.
In the next section we will show that $\Integer$ is subtype of $\Number$.
This allows programmers to keep using \texttt{integer} values where
\texttt{float} values are expected.

The type $\Nil$ is the type of \texttt{nil}, the value that Lua uses for
undefined variables, missing parameters, and missing table keys.

The type $\Value$ is the top type, which represents any Lua value.
In Section \ref{sec:rules} we will show that this type,
along with variadic types, helps the type system to drop extra values
on assignments and function calls, thus preserving the
semantics of Lua in these cases.

Typed Lua uses the type $\Self$ to represent the \emph{receiver}
in object-oriented method definitions and method calls.
As we mentioned in Section \ref{sec:oop}, we need the type
$\Self$ to prevent programs from indexing a method without
calling it with the correct receiver.

@@

Function types have the form $s \rightarrow s$ and represent Lua functions,
where $s$ is a second-level type.

Second-level types are tuples of first-level types that can end
with either an empty tuple or with a variadic type, and
unions of tuple types.
Typed Lua needs second-level types because tuples are not first-class
values in Lua, only appearing on argument passing, multiple returns,
and multiple assignments.
The type $\Void$ is the type of an empty tuple.
A variadic type $t*$ represents a sequence of values of type $t \cup \Nil$;
it is the type of a vararg expression.
Second-level types include unions of tuples because Lua programs
usually overload the return type of functions to denote error,
as we mentioned in Section \ref{sec:statistics}.
For clarity, we use the symbol $\sqcup$ to represent the union between
two different tuple types.
Note that $\cup$ represents the union between two first-level types,
while $\sqcup$ represents the union between two tuple types.

Back to first-level types, table types represent the various forms
that Lua tables can take.
The syntactical form of table types is $\{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{tag}$,
where each $k_{i}$ represents the type of a table key,
and each $v_{i}$ represents the type of the value that table keys of type $k_{i}$ map to.
Key types can only be literal types, base types, or the top type.
We made this restriction to the type of the keys because the statistics
that we discussed in Section \ref{sec:statistics} showed that most
of the tables are records, lists, and hashes.
The type $\Value$ is an option when we need a loose table type.
For instance, $\{\Value:\Value\}_{closed}$ represents the type of a
table that both indices and values can have any type.
Value types can be any first-level type, and can optionally include
the $\Const$ type to denote immutable values.

We also use the tags \emph{closed}, \emph{open}, and \emph{unique}
to classify table types.
The tag \emph{unique} represents tables that have no keys that do not
inhabit one of the table's key types, and that have no alias.
In particular, the type of the table constructor has this tag.
The tag \emph{open} represents \emph{unique} table types that
have at least one alias.
The tag \emph{closed} represents table types that do not provide
any guarantees about keys with types not listed in the table type.
In particular, in the concrete syntax, type annotations, interface
declarations, and userdata declarations always describe \emph{closed} table types.
In the next sections we explain in more detail why we need
different table types.

@@

Typed Lua includes projection types as a way to project
unions of tuple types into unions of first-level types,
as we mentioned in Section \ref{sec:unions}.
In Section \ref{sec:rules} we show in more detail how our type system
uses them as a mechanism for handling unions of tuple types,
when they appear on the right-hand side of the declaration of local variables.
We also show how this feature allows our type system to constrain
the type of a local variable that depends on the type of another local variable.

@@

However, sometimes we want to use \emph{unique} table types to initialize
\emph{closed} table types, as they allow us to initialize optional fields.
To do that, Typed Lua includes a cast expression that allow us
to use an \emph{unique} table type before it becomes \emph{open}.
The rule \textsc{T-CAST} describes this behavior:
\[
\begin{array}{c}
\mylabel{T-CAST}\\
\dfrac{\env_{1}(n) \subtype t}
      {\env_{1} \vdash {<}t{>} \; n:t, \env_{1}[n \mapsto t]}
\end{array}
\]
